Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-05-10 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Comments Valentijn

In this section we'll go over the different steps involved with imputing incomplete multilevel data with the R package mice. We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$). The following variables are of primary interest:

@@

@@

```{r echo=FALSE}
fits <-
  purrr::map_dfr(
    unique(popcomp$school),
    ~ {
      lm(
        mod,
        data = filter(popcomp, school == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(school = as.character(.x), .)
  ) %>% filter(term %in% "sex")

@@

Plot the missing data pattern:

@@

@@

@@

In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels. When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Audigier, 2021), e.g. if the main model accounts for the hierarchical structure also imputation model should do it. Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.

>>>>

In this section we will go over the different steps involved with imputing incomplete multilevel data with the R package mice. We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$). The following variables are of primary interest:

@@

<!-- TODO: choose sex or gender -->

@@

<!-- TODO: True effect? We have not discussed any effects yet. Maybe say that "we consider the associations in the full data set to be the true associations''. And mention association instead of effect (effect has a causal interpretation, association can be either causal or predictive) -->

```{r echo=FALSE, fig.cap = "True effect in the popularity data"}
fits <-
  purrr::map_dfr(
    unique(popcomp$school),
    ~ {
      lm(
        mod,
        data = filter(popcomp, school == .x)
      )
    } %>%
      broom::tidy(conf.int = TRUE, exponentiate = TRUE) %>%
      .[, c("term", "estimate", "conf.low", "conf.high")] %>%
      cbind(school = as.character(.x), .)
  ) %>% filter(term %in% "sex")

@@

First we plot the pattern of missing data within categories of the relevant variables. Plot the missing data pattern:

@@

Print the estimates:

<!-- TODO: explain what this function does -->

```{r}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

@@

- FIML vs MI

An alternative approach to missing data is to use Full Information Maximum Likelihood (FIML). This method does not require the imputation of any missing values. Whereas MI consists of  imputation, analyses and pooling steps, FIML analyses the data in a single step. When the assumptions are met the two approaches should produce equivalent results. [REF] As FIML requires specialised software, not all analyses can be performed with standard software. [REF]

- Survival / TTE, this could be put in the paragraph on congeniality

When the outcome is time-to-event, the Nelson-Aalen estimate of the time to event should be included as a covariate in the imputation model [REF]

@@

In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels. When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g. if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021). Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.
